Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2014-07-16 draft.tex
Rework secion on detector acceptance effects

\fred{Comment that in this section $\vec\nu$ and $\vec\theta$ are used after ``correction'' has been applied.}
\danny{Rethink order of this section.}
Thanks to accurate detector simulations and control studies, the experimental analyses have become
remarkably good at removing detector acceptance effects. In this section we therefore assume that
there are only a small, residual acceptance effects left that affect the events. In particular, we
assume that the latter can be modeled as a modified probability density,
\begin{equation}
    \label{eq:PDFwithDE}
    P(\nuvec,\thvec) \to P(\nuvec,\thvec) \times \Omega(\thvec),
\end{equation}
where $\Omega$ is independent of the non-angular kinematics $\nuvec$. The latter assumption is only
is invalid in general. However, if residual angular acceptance only weakly depends on $\vec{\nu}$, one
can always choose appropriate bin $\vec{\nu}$ so that $\Omega$ is approximately constant with respect
to $\vec{\nu}$. For such a case we parametrize
\begin{equation}
    \Omega_{m,l}(\thvec) = \prod_{i=1}^{m} \Big[1 + \sum_{j=1}^l \eps_i^{(j)} p_j(\cos\theta_i)\big]
\end{equation}
where $p_j(\cos\theta_i)$ denotes the $j$th Legendre polynomial. Since we only consider residual
acceptance effects, we expect $\eps_i^{(j)} \ll 1$. \footnote{\danny{what are reasonable values/upper limits on the $\eps$?}}\\

@@

For this type of distribution, we expand $\Omega_3$ to linear order in $\eps$,
\begin{equation}
    \Omega_{3,2}(\theta_1,\theta_2,\theta_3) = 1 + \sum_{i=1}^3 \sum_{j=1}^2 \eps_i^{(j)} p_j(\cos\theta_i) + \order{\eps^2}\,.
\end{equation}
This leads to a matrix-valued equation that relates the effiency-corrected estimators $\wwhat{S}_i$ to the uncorrected
estimators $\est{S}_j$,
\begin{equation}
    \label{eq:eff-corrected-est-btokstarll}
    \wwhat{S}_i = \sum_{j=1\text{s},\dots\,9} \big(\delta_{ij} + N_{ij}\big) \est{S}_j\,.
\end{equation}
We provide the matrix $N$ in \reftab{eff-corrected-est-btokstarll}.\\

The determination of detector acceptance is generally a difficult task. Here, we will show a systematic
method to its effects.\\

As previous, we use a PDF $P$ decomposed into $N$ angular observables $S_k$. We denote the estimator of the observable
$S_k$ as $\est{S_k}$, which is affected by the detector acceptance. We assume now that the acceptance-corrected estimator
$\wwhat{S_k}$ is a linear combination of the uncorrected estimators $\lbrace \est{S_k}\rbrace$,
\begin{equation}
    \label{eq:linear-assumption}
    \wwhat{S_k} = \sum_j M_{kj} \est{S}_j\,.
\end{equation}
The coefficients $M_{kj}$ can now be systematically determined from MC studies of the detector geometry.
\begin{enumerate}
    \item In step $i$, let $S_i = 1$, and $S_{k\neq i} = 0$.
    \item Simulate events from the PDF, and adjust each event based on the detector geometry. This allows
        to compute $\est{S_j^{(i)}}$ from the simulation.
    \item Solve the system of linear equations
        \begin{equation}
        \begin{aligned}
            1 & = \sum_j M_{ij}^{(i)} \est{S_j^{(i)}}\,,\\
            0 & = \sum_j M_{kj}^{(i)} \est{S_j^{(i)}}\,,\qquad\forall k \neq i\,.
        \end{aligned}
        \end{equation}
        This allows the dermination of those coefficients $M_{kj}^{(i)}$ for which $\est{S_j^{(i)}} \neq 0$.
    \item If $i < N$ advance $i$ by one and go to step 1
\end{enumerate}
We can now check if the assumption \refeq{linear-assumption} is valid, by checking that
$M_{kj}^{(i)} = M_{kj}^{(i')}$ for all $i$, $i'$.

>>>>

The determination of detector acceptance is generally a difficult task. Here, we will show a systematic
method to estimate its effects based on both the method of moments, and MC simulations of the detector.\\

As previous, we use a PDF $P$ decomposed into $N$ angular observables $\lbrace S_k\rbrace$,
\begin{equation}
    P \equiv P(\vec\nu,\vec\theta | \lbrace S_k \rbrace)\,.
\end{equation}
We denote the estimator of an
observable $S_k$ as $\est{S_k}$, which is affected by the detector acceptance. We assume now that there is a linear relation
between the true angular observables $\lbrace S_k^\text{true}\rbrace$, and the estimators $\lbrace \est{S_k}\rbrace$,
\begin{equation}
    \label{eq:linear-assumption}
    \est{S_i} = \sum_{i = 1}^N M_{i j} S_j^\text{true}\,.
\end{equation}
In that case, the matrix $M_{ij}$ can be fully determined through the following steps.
\begin{enumerate}
    \item For a fixed $j$, let $S_j^\text{true} = 1$, and $S_k^\text{true} = 0\,\forall k\neq j$.
    \item Draw simulated events $\lbrace x_t^{(k)}\rbrace = \lbrace(\vec{\nu}^{(k)}_t,\vec\theta^{(k)}_t)\rbrace$ (dubbed ``true'') from the PDF $P_j$,
        \begin{equation}
            P_j \equiv P(\vec\nu,\vec\theta | \lbrace S_j = 1, S_{k\neq j} = 0\rbrace)\,.
        \end{equation}
    \item Use the true events as input for the MC simulation of the detector, in order to obtain the detector
        events $\lbrace x^{(k)}_d\rbrace = \lbrace(\vec{\nu}^{(k)}_d,\vec\theta^{(k)}_d)\rbrace$.
    \item Use the method of moments to determine the uncorrected observables $\lbrace \est{S_k} \rbrace$, based on the detector
        events $\lbrace x^{(k)}_d\rbrace$.
    \item By virtue of \refeq{linear-assumption} and step 1, the $j$-th column of $M$ has been determined via
        \begin{equation}
            M_{ij} = \est{S_i}\,.
        \end{equation}
    \item Repeat the above steps for all values of $j$ to determine all columns of $M$.
\end{enumerate}

It remains to invert the relation \refeq{linear-assumption}, so that one can determine the unfolded
angular observables $\lbrace \wwhat{S_k}\rbrace$ as
\begin{equation}
    \wwhat{S_i} = \left(M^{-1}\right)_{ij} \est{S_j}\,.
\end{equation}

We emphasize that this methods hinges on the assumption in \refeq{linear-assumption}, which might not be fulfilled in general.
In particular, if detector effects are small, of it they can be reduced through other means, our approach is expected to work.
The effects can be considered small if
\begin{equation}
    ||M - \mathbf{1}||_2 \ll N\,.
\end{equation}
Here $||\cdot||_2$ denotes the Frobenius norm.

\danny{$@$Marcin: How well does this work? Is \refeq{linear-assumption} fulfilled in your cases?}

\vspace{\bigskipamount}
\hrule
\danny{Should we remove the remainder of this section?}
\fred{Comment that in this section $\vec\nu$ and $\vec\theta$ are used after ``correction'' has been applied.}
Thanks to accurate detector simulations and control studies, the experimental analyses have become
remarkably good at removing detector acceptance effects. In this section we therefore assume that
there are only a small, residual acceptance effects left that affect the events. In particular, we
assume that the latter can be modeled as a modified probability density,
\begin{equation}
    \label{eq:PDFwithDE}
    P(\nuvec,\thvec) \to P(\nuvec,\thvec) \times \Omega(\thvec),
\end{equation}
where $\Omega$ is independent of the non-angular kinematics $\nuvec$. The latter assumption is only
is invalid in general. However, if residual angular acceptance only weakly depends on $\vec{\nu}$, one
can always choose appropriate bin $\vec{\nu}$ so that $\Omega$ is approximately constant with respect
to $\vec{\nu}$. For such a case we parametrize
\begin{equation}
    \Omega_{m,l}(\thvec) = \prod_{i=1}^{m} \Big[1 + \sum_{j=1}^l \eps_i^{(j)} p_j(\cos\theta_i)\big]
\end{equation}
where $p_j(\cos\theta_i)$ denotes the $j$th Legendre polynomial. Since we only consider residual
acceptance effects, we expect $\eps_i^{(j)} \ll 1$. \footnote{\danny{what are reasonable values/upper limits on the $\eps$?}}\\

@@

For this type of distribution, we expand $\Omega_3$ to linear order in $\eps$,
\begin{equation}
    \Omega_{3,2}(\theta_1,\theta_2,\theta_3) = 1 + \sum_{i=1}^3 \sum_{j=1}^2 \eps_i^{(j)} p_j(\cos\theta_i) + \order{\eps^2}\,.
\end{equation}
This leads to a matrix-valued equation that relates the effiency-corrected estimators $\wwhat{S}_i$ to the uncorrected
estimators $\est{S}_j$,
\begin{equation}
    \label{eq:eff-corrected-est-btokstarll}
    \wwhat{S}_i = \sum_{j=1\text{s},\dots\,9} \big(\delta_{ij} + N_{ij}\big) \est{S}_j\,.
\end{equation}
We provide the matrix $N$ in \reftab{eff-corrected-est-btokstarll}.\\
\hrule
